\large \textbf{Leveraging Historical Ties between Cognitive Science and Computer Science to Guide Programming Education}

### 2016-02-04 paper.tex
added example patterns for procedural and oop

%% TODO: more directed title
%% TODO: possibly use subheadings to ``tweet'' the major takeaways from each section,
%% this could then be connected at the end to bullet points of suggestions
\section{Trends in programming education research}
A \citeyear{robins_learning_2003} study by
\citeauthor{robins_learning_2003} identifies four trends in research
of programming education: 1) distinguishing between novice and expert
programmers with emphasis on deficiencies of novices, 2) the
distinction between knowledge and strategies, 3) the distinction
between program comprehension and generation, and 4) a comparison of
OOP to procedural styles of
programming\autocite{robins_learning_2003}. They found that the
biggest challenge for novice programmers was not understanding the
basic concepts, but knowing how to apply them. This finding is
consistent understanding of how we
learn\autocite{bransford_how_2000,ambrose_chapter_2010}, and with
studies of expert/novice behavior in other domains, such as
statics\autocite{litzinger_cognitive_2010}. In both cases the
suggestion is that helping students learn strategies that work is
crucial to helping them become effective novices, i.e. devote
instructional time to problem-solving strategies and how to apply
them.

@@

There remains a desire that introductory programming courses should
teach problem-solving skills, though only a few papers present
evidence that the implementation of this goal is informed by cognitive
science\autocite{kramarski_cognitivemetacognitive_1997,reed_incorporating_1998,muller_supporting_2008}. One
thread in particular has received regular attention:
\citeauthor{east_patternbased_1996} propose a pattern-based approach
that places emphasis on recognizing and learning to work with
algorithmic patterns over syntax and design
paradigms\autocite{east_patternbased_1996}. This approach closely
aligns with understanding of differences in expert/novice behavior by
helping students identify patterns of solutions and learning how and
when to apply them to particular
problems\autocite{east_patternbased_1996,muller_pattern_2005,wallingford_first_1996}.

@@

Alan Kay originally conceived of object oriented programming as
\emph{one} way to think about complex
systems\autocite{kay_history_1996}. The underlying metaphore of Kay's
conception of OOP was that of the biological cell: each object is
designed to accomplish a specific set of tasks where the complexity of
these tasks is encapsulated within the object. Objects interact with
one another by sending and receiving messages, and systems of these
objects are assembled to solve complex
problems\autocite{kay_history_1996}.

@@

In several reports of attempts to incorporate OOP into introductory
programming courses, practitioners worried that time devoted to OOP
concepts would necessarily reduce time spent on the ``basics of
programming'', usually defined as conditional statements, loops, and
in some cases pointers. This concern is misplaced, under an OOP
paradigm the ``basics of programming'' are not the same as when
working from a procedural paradigm. OOP defines a completely different
set of tools for problem solving, and it is these tools and concepts
that become the
``basics''\autocite{kolling_problem_1999,kay_history_1996}.

For instance, if taking a patterns-based approach to teaching
programming one could use either a procedural paradigm or an OOP
paradigm, but the fundamental patterns would be quite different
depending on the paradigm\autocite{wallingford_first_1996}.

>>>>

%% TODO: more directed title
%% TODO: possibly use subheadings to ``tweet'' the major takeaways from each section,
%% this could then be connected at the end to bullet points of suggestions
\section{Trends in programming education research}
A \citeyear{robins_learning_2003} study by
\citeauthor{robins_learning_2003} identifies four trends in research
of programming education: 1) distinguishing between novice and expert
programmers with emphasis on deficiencies of novices, 2) the
distinction between knowledge and strategies, 3) the distinction
between program comprehension and generation, and 4) a comparison of
OOP to procedural styles of
programming\autocite{robins_learning_2003}. They found that the
biggest challenge for novice programmers was not in understanding the
basic concepts, but knowing how to apply them. This finding is
consistent with current understanding of how we
learn\autocite{bransford_how_2000,ambrose_chapter_2010}, and with
studies of expert/novice behavior in other domains, such as
statics\autocite{litzinger_cognitive_2010}. In both cases the
suggestion is that helping students learn strategies that work is
crucial to helping them become \emph{effective} novices, i.e. devote
instructional time to problem-solving strategies and how to apply
them.

@@

There remains a desire that introductory programming courses should
teach problem-solving skills, though only a few papers present
evidence that the implementation of this goal is informed by cognitive
science\autocite{kramarski_cognitivemetacognitive_1997,reed_incorporating_1998,muller_supporting_2008}. One
thread in particular has received regular attention:
\citeauthor{east_patternbased_1996} propose a pattern-based approach
that places emphasis on recognizing and learning to work with
algorithmic patterns over syntax and design
paradigms\autocite{east_patternbased_1996}. This approach closely
aligns with understanding of differences in expert/novice behavior by
helping students identify patterns of solutions and learning how and
when to apply them to particular
problems\autocite{east_patternbased_1996,muller_pattern_2005,wallingford_first_1996,reed_incorporating_1998}.

@@

Alan Kay originally conceived of object oriented programming as
\emph{one} way to think about complex
systems\autocite{kay_history_1996}. The underlying metaphore of Kay's
conception for OOP was the biological cell: each object is
designed to accomplish a specific set of tasks where the complexity of
these tasks is encapsulated within the object. Objects interact with
one another by sending and receiving messages, and systems of these
objects are assembled to solve complex
problems\autocite{kay_history_1996}.

@@

In several reports of attempts to incorporate OOP into introductory
programming courses, practitioners worried that time devoted to OOP
concepts would necessarily reduce time spent on the ``basics of
programming'', usually defined as conditional statements, loops, and
in some cases pointers. This concern is misplaced, under an OOP
paradigm the ``basics of programming'' are not the same as when
working from a procedural paradigm. OOP defines a completely different
set of tools for problem solving, and it is these tools and concepts
that become the ``basics'': objects, messages, inheritance, and
polymorphism\autocite{kolling_problem_1999,kay_history_1996}.

For instance, if taking a patterns-based approach to teaching
programming, one could use either a procedural paradigm or an OOP
paradigm, but the fundamental patterns would be quite different
depending on the paradigm\autocite{wallingford_first_1996}. Useful
procedural patterns might be:

\begin{description}
\item[read-evaluate-print] reading a piece of data, evaluating it, and output the result
\item[guarded-action] if a guard-condition is satisfied, take action.
\end{description}

While useful OOP patterns might be:

\begin{description}
\item[state]  provide controlled access to a body of data
\item[view] decouple an object's state from its representation
\item[decorator] add functionally to an object without modifying its internal structure.
\end{description}
